### 2020-10-14 priority-gv.tex
Simon's comments.

GV enforces deadlock freedom with a syntactic restriction: it fuses channel creation and thread spawning into a single function, called fork. This guarantees that if some term $\tm{L}$ has the means to communicate with both $\tm{M}$ and $\tm{N}$, then $\tm{M}$ and $\tm{N}$ cannot communicate amongst themselves. In other words, GV only allows tree-shaped communication structures, so it is trivially free from cycles, and therefore deadlocks. However, this is overly restrictive, as it rules out \emph{all} cyclic communication structures, not just those with deadlocks.

Priority GV offers a more fine-grained analysis of the communication structure, and allows cyclic communication structures. To illustrate this, we use PGV to implement a variant of Milner's cyclic scheduler in~\cref{sec:pgv-example}.

\subsubsection*{Types}
Types in PGV are divided into session types ($\ty{S}$) and types ($\ty{T}$, $\ty{U}$).
%
Session types are defined by the following grammar:
\[
\begin{array}{lcl}
  \ty{S}
  & \coloneqq & \ty{\tysend[\cs{o}]{T}{S}}
    \sep        \ty{\tyrecv[\cs{o}]{T}{S}}
    \sep        \ty{\tyends[\cs{o}]}
    \sep        \ty{\tyendr[\cs{o}]}
\end{array}
\]
The session types $\ty{\tysend[\cs{o}]{T}{S}}$ and $\ty{\tyrecv[\cs{o}]{T}{S}}$ describe the endpoints of a channel over which we send or receive a value of type $\ty{T}$, and then proceed as $\ty{S}$. The types $\ty{\tyends[\cs{o}]}$ and $\ty{\tyendr[\cs{o}]}$ describe endpoints of a channel whose communication has finished, and over which we must synchronise before closing the channel. Each connective in a session type is annotated with a \emph{priority} $\cs{o}\in\mathbb{N}$.

Types are defined by the following grammar:
\[
\begin{array}{lcl}
  \ty{T}, \ty{U}
  & \coloneqq & \ty{\typrod{T}{U}}
    \sep        \ty{\tyunit}
    \sep        \ty{\tysum{T}{U}}
    \sep        \ty{\tyvoid}
    \sep        \ty{\tylolli[\cs{p},\cs{q}]{T}{U}}
    \sep        \ty{S}
\end{array}
\]
The types $\ty{\typrod{T}{U}}$, $\ty{\tyunit}$, $\ty{\tysum{T}{U}}$, and $\ty{\tyvoid}$ are the standard linear $\lambda$-calculus product type, unit type, sum type, and empty type.
The type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$ is the standard linear function type, annotated with \emph{priority bounds} $\cs{p},\cs{q}\in\mathbb{N}\cup\{\cs{\pbot},\cs{\ptop}\}$.
Every session type is also a type.

@@

\subsubsection*{Environments}
Typing environments $\ty{\Gamma}$, $\ty{\Delta}$ associate names to types. Environments are linear, so two environments can only be combined as $\ty{\Gamma}, \ty{\Delta}$ if their names are distinct, \ie $\fv(\ty{\Gamma})\cap\fv(\ty{\Delta})=\varnothing$.
\[
\begin{array}{lcl}
  \ty{\Gamma}, \ty{\Delta}
  & \coloneqq & \ty{\emptyenv}
    \sep        \ty{\Gamma}, \tmty{x}{T}
\end{array}
\]

@@

\subsubsection*{Priorities}
The function $\pr$ returns the smallest priority of a session type. The type system guarantees that the top-most connective always holds the smallest priority, so we simply return the priority of the top-most connective:
\[
\begin{array}{lcl}
  \pr(\ty{\tysend[\cs{o}]{T}{S}})            & = & \cs{o} \\
  \pr(\ty{\tyrecv[\cs{o}]{T}{S}})            & = & \cs{o} \\
\end{array}
\qquad
\begin{array}{lcl}
  \pr(\ty{\tyends[\cs{o}]})                  & = & \cs{o} \\
  \pr(\ty{\tyendr[\cs{o}]})                  & = & \cs{o} \\
\end{array}
\]
We extend the function $\pr$ to types and typing contexts by returning the smallest priority in the type or context, or $\cs{\top}$ if there is no priority:
\[
\begin{array}{lcl}
  \minpr(\ty{\typrod{T}{U}})                 & = & \minpr({\ty{T}})\sqcap\minpr({\ty{U}}) \\
  \minpr(\ty{\tysum{T}{U}})                  & = & \minpr({\ty{T}})\sqcap\minpr({\ty{U}}) \\
  \minpr(\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}) & = & \cs{p} \\
  \minpr(\ty{\Gamma}, \tmty{x}{A})           & = & \minpr(\ty{\Gamma})\sqcap\pr(\ty{A})
\end{array}
\qquad
\begin{array}{lcl}
  \minpr(\ty{\tyunit})                       & = & \ptop \\
  \minpr(\ty{\tyvoid})                       & = & \ptop \\
  \minpr(\ty{S})                             & = & \pr(\ty{S}) \\
  \minpr(\ty{\emptyenv})                     & = & \ptop
\end{array}
\]

\subsubsection*{Terms}
Terms ($\tm{L}$, $\tm{M}$, $\tm{N}$) are defined by the following grammar:
\[
\begin{array}{lcl}
  \tm{L}, \tm{M}, \tm{N}
  & \coloneqq & \tm{x}
    \sep        \tm{K}
    \sep        \tm{\lambda x.M}
    \sep        \tm{M\;N} \\
  & \sep      & \tm{\unit}
    \sep        \tm{\letunit{M}{N}} \\
  & \sep      & \tm{\pair{M}{N}}
    \sep        \tm{\letpair{x}{y}{M}{N}} \\
  & \sep      & \tm{\inl{M}}
    \sep        \tm{\inr{M}}
    \sep        \tm{\casesum{L}{x}{M}{y}{N}} \\
  & \sep      & \tm{\absurd{M}} \\
  \tm{K}
  & \coloneqq & \tm{\link}
    \sep        \tm{\new}
    \sep        \tm{\spawn}
    \sep        \tm{\send}
    \sep        \tm{\recv}
    \sep        \tm{\close}
    \sep        \tm{\wait}
\end{array}
\]
Let $\tm{x}$, $\tm{y}$, $\tm{z}$, and $\tm{w}$ range over variable names. Occasionally, we use $\tm{a}$, $\tm{b}$, $\tm{c}$, and $\tm{d}$, which range over \emph{free} endpoints. The term language is the standard linear $\lambda$-calculus with products, sums, and their units, extended with constants $\tm{K}$ for the communication primitives.

The constants are best understood in conjunction with their typing and reduction rules in~\cref{fig:pgv-typing,fig:pgv-operational-semantics}. Briefly, $\tm{\link}$ links two endpoints together, forwarding messages from each to the other, $\tm{\new}$ creates a new channel and returns its endpoints, and $\tm{\spawn}$ spawns off its argument as a new thread. The $\tm{send}$ and $\tm{recv}$ functions send and receive values on a channel. However, since the typing rules for PGV ensure the linear usage of endpoints, they also return a new copy of the endpoint, to continue the session. The $\tm{close}$ and $\tm{wait}$ functions close a channel.

We frequently use syntactic sugar to make terms more readable: we write $\tm{\andthen{M}{N}}$ in place of $\tm{\letunit{M}{N}}$, $\tm{\letbind{x}{M}{N}}$ in place of $\tm{(\lambda x.N)\;M}$, and pattern matching lambdas $\tm{\lambda\unit.M}$ and $\tm{\lambda\pair{x}{y}.M}$ in place of $\tm{\lambda z.\letunit{z}{M}}$ and $\tm{\lambda z.\letpair{x}{y}{z}{M}}$.

@@

\subsubsection*{Operational Semantics}
Priority GV terms are evaluated as part of a configuration of processes. Configurations are defined by the following grammar:
\[
\begin{array}[t]{lcl}
    \tm{\conf{C}}, \tm{\conf{D}}, \tm{\conf{E}}
  & \coloneqq & \tm{\phi\;M}
    \sep        \tm{\ppar{\conf{C}}{\conf{D}}}
    \sep        \tm{\res{x}{x'}{\conf{C}}}
  \\
  \tm{\phi}
  & \coloneqq & \tm{\main}
    \sep        \tm{\child}
\end{array}
\]
Configurations ($\tm{\conf{C}}$, $\tm{\conf{D}}$, $\tm{\conf{E}}$) consist of threads $\tm{\phi\;M}$, parallel compositions $\tm{\ppar{\conf{C}}{\conf{D}}}$, and name restrictions $\tm{\res{x}{x'}{\conf{C}}}$. To preserve the functional nature of PGV, where programs return a single value, we use flags ($\tm{\phi}$) to differentiate between the main thread, marked $\tm{\main}$, and child threads created by $\tm{\spawn}$, marked $\tm{\child}$. Only the main thread returns a value. We determine the flag of a configuration by combining the flags of all threads in that configuration:
\[
\begin{array}{ll}
  \tm{\main}  + \tm{\child} = \tm{\main}
  \\
  \tm{\child} + \tm{\main}  = \tm{\main}
\end{array}
\quad
\begin{array}{ll}
  \tm{\child} + \tm{\child} = \tm{\child}
  \\
  \tm{\main}  + \tm{\main} \; \text{undefined}
\end{array}
\]
Values ($\tm{V}$, $\tm{W}$), ready terms ($\tm{R}$) evaluation contexts ($\tm{E}$), thread evaluation contexts ($\tm{\conf{F}}$), and configuration contexts ($\tm{\conf{G}}$) are defined by the following grammars:
\[
\begin{array}[t]{lcl}
  \tm{V}, \tm{W}
  & \coloneqq & \tm{x}
    \sep        \tm{K}
    \sep        \tm{\lambda x.M}
    \sep        \tm{\unit}
    \sep        \tm{\pair{M}{N}}
    \sep        \tm{\inl{M}}
    \sep        \tm{\inr{M}}
  \\
  \tm{E}
  & \coloneqq & \tm{\hole} \\
  & \sep      & \tm{E\;M}
    \sep        \tm{V\;E} \\
  & \sep      & \tm{\letunit{E}{N}} \\
  & \sep      & \tm{\pair{E}{M}}
    \sep        \tm{\pair{V}{E}}
    \sep        \tm{\letpair{x}{y}{E}{M}} \\
  & \sep      & \tm{\inl{E}}
    \sep        \tm{\inr{E}}
    \sep        \tm{\casesum{E}{x}{M}{y}{N}} \\
  & \sep      & \tm{\absurd{E}}
  \\
  \tm{\conf{F}}
  & \coloneqq & \tm{\phi\;E}
  \\
  \tm{\conf{G}}
  & \coloneqq & \tm{\hole}
    \sep        \tm{\ppar{\conf{G}}{\conf{C}}}
    \sep        \tm{\res{x}{y}{\conf{G}}}
\end{array}
\]
We factor the reduction relation of PGV into a deterministic reduction on terms ($\tred$) and a non-deterministic reduction on configurations ($\cred$), see \cref{fig:pgv-operational-semantics}. We write $\tred^+$ and $\cred^+$ for the transitive closures, and $\tred^\star$ and $\cred^\star$ for the reflexive-transitive closures.

@@

The only non-standard feature of the typing rules is the priority annotations. Priorities are based on \emph{obligations/capabilities} used by Kobayashi~\cite{kobayashi06}, and simplified to single priorities following Padovani~\cite{padovani14}. The integration of priorities into GV is adapted from Padovani and Novara~\cite{padovaninovara15}. Paraphrasing from Dardha and Gay~\cite{dardhagay18}, priorities obey the following two laws:
\begin{enumerate}
  \item an action with priority $\cs{o}$ happens before an action with priority $\cs{o+n}$;
  \item communication requires \emph{equal} priorities for the dual actions.
\end{enumerate}

In PGV, Sequents are annotated by a priority bound $\cs{p}$, which is the \emph{upper bound} on the priority of the communication actions performed while evaluating the term. The \emph{lower bound} is obtained from the typing environment. Terms with sequential evaluation, such as \LabTirName{T-LetUnit}, have side conditions to ensure deadlock freedom. For instance, the side condition for typing $\tm{\letunit{M}{N}}$ requires that $\tm{M}$ has finished communicating before $\tm{N}$ starts communicating.

Configurations are typed by a judgement $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$, which states that ``a configuration $\tm{\conf{C}}$ with flag $\phi$ is well-typed under typing environment $\ty{\Gamma}$''. Configuration typing is based on the standard typing for GV. Terms are embedded either as main or as child threads. The priority bound from the term typing is discarded, as configurations contain no further sequential actions. Main threads are allowed to return a value, whereas child threads are required to return the unit value. Sequents are annotated with a flag $\phi$, which ensure that there is at most one main thread.

@@

Secondly, we simplify typing for parallel composition. In GV, each parallel composition must split \emph{exactly one} channel of the channel pseudo-type $\ty{S^\sharp}$ into two endpoints of type $\ty{S}$ and $\ty{\co{S}}$ across the parallel composition. Consequently, associativity of parallel composition violates type preservation. Unfortunately, the practice plays a crucial role in guaranteeing deadlock freedom. In PGV, we guarantee deadlock freedom through different means, which removes the channel pseudo-type $\ty{S^\sharp}$, simplifies typing for parallel composition, and restores type preservation for the structural congruence.

@@

\subsubsection*{Example: Milner's Cyclic Scheduler}\label{sec:pgv-example}
We illustrate the newfound ability of PGV by encoding a variant of Milner's cyclic scheduler~\cite{milner89,dardhagay18}. Suppose a set of processes $\tm{\proc{i}}$, $1\leq{i}\leq{n}$, should be scheduled to perform some tasks in cyclic order, starting with $\tm{\proc0}$, and circling all the way around to $\tm{\proc{n}}$, notifying $\tm{\proc0}$ when every other process has finished their task.

>>>>

GV enforces deadlock freedom with a syntactic restriction: it fuses channel creation and thread spawning into a single function, called fork\simon{It feels like you've said this a *lot* at this point}. This guarantees that if some term $\tm{L}$ has the means to communicate with both $\tm{M}$ and $\tm{N}$, then $\tm{M}$ and $\tm{N}$ cannot communicate amongst themselves. In other words, GV only allows tree-shaped communication structures, so it is trivially free from cycles, and therefore deadlocks. However, this is overly restrictive, as it rules out \emph{all} cyclic communication structures, not just those with deadlocks.

Priority GV offers a more fine-grained analysis of the communication structure, and allows cyclic communication structures. To illustrate this, we start by examining a variant of Milner's cyclic scheduler in PGV~\cite{milner89}.

\subsubsection*{Example}\label{sec:pgv-example}
Suppose a set of processes $\tm{\proc{i}}$, $1\leq{i}\leq{n}$, should be scheduled to perform some tasks in cyclic order, starting with $\tm{\proc0}$, and circling all the way around to $\tm{\proc{n}}$, notifying $\tm{\proc0}$ when every other process has finished their task.

Our scheduler consists of set of agents $\tm{\agent{i}}$, $1\leq{i}\leq{n}$, each representing their respective process. Each process $\tm{\proc{i}}$ waits for the signal to start their task on $\tm{a'_i}$, and signals completion on $\tm{b'_i}$. Each agent signals their process to start on $\tm{a_i}$, waits for their process to finish on $\tm{b_i}$, and then signals for the next agent to continue on $\tm{c_i}$. The agent $\tm{\agent0}$ initiates, then waits for every other process to finish, and signals $\tm{\proc0}$ on $\tm{d}$. Every other agent $\tm{\agent{i}}$, $2\leq{i}\leq{n}$ waits on $\tm{c'_{i-1}}$ for the signal to start. The dotted lines represent the optional transfer of data between processes. With three processes, the scheduler looks as follows:\simon{Sorry, I found this diagram super confusing\dots Is there any way you could make it look more cyclic? Also, have the convention of a single annotation per channel, with the convention that the greater agent / proc gets the primed end of the previous channel?}
\input{fig/pgv-scheduler.tex}
We implement the scheduler as follows, using $\tm{\prod_{i}\conf{C}_i}$ to denote the parallel composition of the processes $\tm{\conf{C}_i}$, and $\tm{\plug{M}{N}}$ to denote the plugging of $\tm{N}$ in the one-hole term context $\tm{M}$:
\[
  \begin{array}{lrlr}
    \tm{\sched}
    & \defeq & \tm{\res{a_1}{a'_1}{\dots{\res{a_n}{a'_n}{}}}}
               \tm{\res{b_1}{b'_1}{\dots{\res{b_n}{b'_n}{}}}}
               \tm{\res{c_1}{c'_1}{\dots{\res{c_n}{c'_n}{}}}}
               \tm{\res{d}{d'}{}}
    \\ &     & \tm{\begin{array}{lll}
                     (
                     & \phi\;\proc1
                     \parallel
                     \child\;\andthen{\agent1}{\andthen{\wait\;{c'_3}}{\close\;{d}}}
                     \\
                     \parallel
                     &
                     \prod_{2\leq{i}\leq{n}}
                     (\child\;\proc{i} \parallel \child\;\andthen{\wait\;{c'_{i-1}}}{\agent{i}})
                     & )
                   \end{array}}
    \\
    \tm{\agent{i}}
    & \defeq & \tm{\andthen{\close\;{a_i}}{\andthen{\wait\;{b_i}}{\close\;{c_i}}}}
    \\
    \tm{\proc{i}}
    & \defeq & \tm{\andthen{\wait\;{a'_i}}{\plug{M_i}{\andthen{\close\;{b'_i}}{N_i}}}}
  \end{array}
\]
Each of the channels in the scheduler is has the terminated type, and is merely used to synchronise.
For simplicity, we let $\tm{\sched}$ be a configuration. However, we could equivalently have represented it as a function which takes three processes as arguments, and creates this configuration.

\subsubsection*{Types}
Types in PGV are divided into session types ($\ty{S}$) and types ($\ty{T}$, $\ty{U}$).
%
Session types are defined by the following grammar:
\[
\begin{array}{lcl}
  \ty{S}
  & \Coloneqq & \ty{\tysend[\cs{o}]{T}{S}}
    \sep        \ty{\tyrecv[\cs{o}]{T}{S}}
    \sep        \ty{\tyends[\cs{o}]}
    \sep        \ty{\tyendr[\cs{o}]}
\end{array}
\]
The session types $\ty{\tysend[\cs{o}]{T}{S}}$ and $\ty{\tyrecv[\cs{o}]{T}{S}}$ describe the endpoints of a channel over which we send or receive a value of type $\ty{T}$, and then proceed as $\ty{S}$. The types $\ty{\tyends[\cs{o}]}$ and $\ty{\tyendr[\cs{o}]}$ describe endpoints of a channel whose communication has finished, and over which we must synchronise before closing the channel. Each connective in a session type is annotated with a \emph{priority} $\cs{o}\in\mathbb{N}$.

Types are defined by the following grammar:
\[
\begin{array}{lcl}
  \ty{T}, \ty{U}
  & \Coloneqq & \ty{\typrod{T}{U}}
    \sep        \ty{\tyunit}
    \sep        \ty{\tysum{T}{U}}
    \sep        \ty{\tyvoid}
    \sep        \ty{\tylolli[\cs{p},\cs{q}]{T}{U}}
    \sep        \ty{S}
\end{array}
\]
The types $\ty{\typrod{T}{U}}$, $\ty{\tyunit}$, $\ty{\tysum{T}{U}}$, and $\ty{\tyvoid}$ are the standard linear $\lambda$-calculus product type, unit type, sum type, and empty type.
The type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$ is the standard linear function type, annotated with \emph{priority bounds} $\cs{p},\cs{q}\in\mathbb{N}\cup\{\cs{\pbot},\cs{\ptop}\}$.
Every session type is also a type.

@@

\subsubsection*{Environments}
Typing environments $\ty{\Gamma}$, $\ty{\Delta}$ associate names to types. Environments are linear, so two environments can only be combined as $\ty{\Gamma}, \ty{\Delta}$ if their names are distinct, \ie $\fv(\ty{\Gamma})\cap\fv(\ty{\Delta})=\varnothing$.
\[
\begin{array}{lcl}
  \ty{\Gamma}, \ty{\Delta}
  & \Coloneqq & \ty{\emptyenv}
    \sep        \ty{\Gamma}, \tmty{x}{T}
\end{array}
\]

@@

\subsubsection*{Priorities}
The function $\pr$ returns the smallest priority of a session type. The type system guarantees that the top-most connective always holds the smallest priority, so we simply return the priority of the top-most connective:
\[
\begin{array}{lcl}
  \pr(\ty{\tysend[\cs{o}]{T}{S}})            & = & \cs{o} \\
  \pr(\ty{\tyrecv[\cs{o}]{T}{S}})            & = & \cs{o} \\
\end{array}
\qquad
\begin{array}{lcl}
  \pr(\ty{\tyends[\cs{o}]})                  & = & \cs{o} \\
  \pr(\ty{\tyendr[\cs{o}]})                  & = & \cs{o} \\
\end{array}
\]
We extend the function $\pr$ to types and typing contexts by returning the smallest priority in the type or context, or $\cs{\top}$ if there is no priority. We use $\sqcap$ and $\sqcup$ to denote the minimum and maximum:
\[
\begin{array}{lcl}
  \minpr(\ty{\typrod{T}{U}})                 & = & \minpr({\ty{T}})\sqcap\minpr({\ty{U}}) \\
  \minpr(\ty{\tysum{T}{U}})                  & = & \minpr({\ty{T}})\sqcap\minpr({\ty{U}}) \\
  \minpr(\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}) & = & \cs{p} \\
  \minpr(\ty{\Gamma}, \tmty{x}{A})           & = & \minpr(\ty{\Gamma})\sqcap\pr(\ty{A})
\end{array}
\qquad
\begin{array}{lcl}
  \minpr(\ty{\tyunit})                       & = & \ptop \\
  \minpr(\ty{\tyvoid})                       & = & \ptop \\
  \minpr(\ty{S})                             & = & \pr(\ty{S}) \\
  \minpr(\ty{\emptyenv})                     & = & \ptop
\end{array}
\]

\subsubsection*{Terms}
Terms ($\tm{L}$, $\tm{M}$, $\tm{N}$) are defined by the following grammar:
\[
\begin{array}{lcl}
  \tm{L}, \tm{M}, \tm{N}
  & \Coloneqq & \tm{x}
    \sep        \tm{K}
    \sep        \tm{\lambda x.M}
    \sep        \tm{M\;N} \\
  & \sep      & \tm{\unit}
    \sep        \tm{\letunit{M}{N}} \\
  & \sep      & \tm{\pair{M}{N}}
    \sep        \tm{\letpair{x}{y}{M}{N}} \\
  & \sep      & \tm{\inl{M}}
    \sep        \tm{\inr{M}}
    \sep        \tm{\casesum{L}{x}{M}{y}{N}} \\
  & \sep      & \tm{\absurd{M}} \\
  \tm{K}
  & \Coloneqq & \tm{\link}
    \sep        \tm{\new}
    \sep        \tm{\spawn}
    \sep        \tm{\send}
    \sep        \tm{\recv}
    \sep        \tm{\close}
    \sep        \tm{\wait}
\end{array}
\]
Let $\tm{x}$, $\tm{y}$, $\tm{z}$, and $\tm{w}$ range over variable names. Occasionally, we use $\tm{a}$, $\tm{b}$, $\tm{c}$, and $\tm{d}$. The term language is the standard linear $\lambda$-calculus with products, sums, and their units, extended with constants $\tm{K}$ for the communication primitives.

The constants are best understood in conjunction with their typing and reduction rules in~\cref{fig:pgv-typing,fig:pgv-operational-semantics}. Briefly, $\tm{\link}$ links two endpoints together, forwarding messages from each to the other, $\tm{\new}$ creates a new channel and returns a pair of its endpoints, and $\tm{\spawn}$ spawns off its argument as a new thread. The $\tm{\send}$ and $\tm{\recv}$ functions send and receive values on a channel. However, since the typing rules for PGV ensure the linear usage of endpoints, they also return a new copy of the endpoint, to continue the session. The $\tm{\close}$ and $\tm{\wait}$ functions close a channel.

We frequently use syntactic sugar to make terms more readable: we write $\tm{\andthen{M}{N}}$ in place of $\tm{\letunit{M}{N}}$, $\tm{\letbind{x}{M}{N}}$ in place of $\tm{(\lambda x.N)\;M}$, and pattern matching lambdas $\tm{\lambda\unit.M}$ and $\tm{\lambda\pair{x}{y}.M}$ in place of $\tm{\lambda z.\letunit{z}{M}}$ and $\tm{\lambda z.\letpair{x}{y}{z}{M}}$. Finally, we recover $\tm{\fork}$:
\[
  \tm{\fork} \defeq \tm{\lambda x.\letpair{y}{z}{\new\;\unit}{\andthen{\spawn\;{(\lambda\unit.x\;y)}}{z}}}
\]

@@

\subsubsection*{Operational Semantics}
Priority GV terms are evaluated as part of a configuration of processes. Configurations are defined by the following grammar:
\[
\begin{array}[t]{lcl}
    \tm{\conf{C}}, \tm{\conf{D}}, \tm{\conf{E}}
  & \Coloneqq & \tm{\phi\;M}
    \sep        \tm{\ppar{\conf{C}}{\conf{D}}}
    \sep        \tm{\res{x}{x'}{\conf{C}}}
  \\
  \tm{\phi}
  & \Coloneqq & \tm{\main}
    \sep        \tm{\child}
\end{array}
\]
Configurations ($\tm{\conf{C}}$, $\tm{\conf{D}}$, $\tm{\conf{E}}$) consist of threads $\tm{\phi\;M}$, parallel compositions $\tm{\ppar{\conf{C}}{\conf{D}}}$, and name restrictions $\tm{\res{x}{x'}{\conf{C}}}$. To preserve the functional nature of PGV, where programs return a single value, we use flags ($\tm{\phi}$) to differentiate between the main thread, marked $\tm{\main}$, and child threads created by $\tm{\spawn}$, marked $\tm{\child}$. Only the main thread returns a value. We determine the flag of a configuration by combining the flags of all threads in that configuration:
\[
\begin{array}{ll}
  \tm{\main}  + \tm{\child} = \tm{\main}
  \\
  \tm{\child} + \tm{\main}  = \tm{\main}
\end{array}
\quad
\begin{array}{ll}
  \tm{\child} + \tm{\child} = \tm{\child}
  \\
  \tm{\main}  + \tm{\main} \; \text{undefined}
\end{array}
\]
Values ($\tm{V}$, $\tm{W}$), evaluation contexts ($\tm{E}$), thread evaluation contexts ($\tm{\conf{F}}$), and configuration contexts ($\tm{\conf{G}}$) are defined by the following grammars:
\[
\begin{array}[t]{lcl}
  \tm{V}, \tm{W}
  & \Coloneqq & \tm{x}
    \sep        \tm{K}
    \sep        \tm{\lambda x.M}
    \sep        \tm{\unit}
    \sep        \tm{\pair{V}{W}}
    \sep        \tm{\inl{V}}
    \sep        \tm{\inr{V}}
  \\
  \tm{E}
  & \Coloneqq & \tm{\hole} \\
  & \sep      & \tm{E\;M}
    \sep        \tm{V\;E} \\
  & \sep      & \tm{\letunit{E}{N}} \\
  & \sep      & \tm{\pair{E}{M}}
    \sep        \tm{\pair{V}{E}}
    \sep        \tm{\letpair{x}{y}{E}{M}} \\
  & \sep      & \tm{\inl{E}}
    \sep        \tm{\inr{E}}
    \sep        \tm{\casesum{E}{x}{M}{y}{N}} \\
  & \sep      & \tm{\absurd{E}}
  \\
  \tm{\conf{F}}
  & \Coloneqq & \tm{\phi\;E}
  \\
  \tm{\conf{G}}
  & \Coloneqq & \tm{\hole}
    \sep        \tm{\ppar{\conf{G}}{\conf{C}}}
    \sep        \tm{\res{x}{y}{\conf{G}}}
\end{array}
\]
We factor the reduction relation of PGV into a deterministic reduction on terms ($\tred$) and a non-deterministic reduction on configurations ($\cred$), see \cref{fig:pgv-operational-semantics}. We write $\tred^+$ and $\cred^+$ for the transitive closures, and $\tred^\star$ and $\cred^\star$ for the reflexive-transitive closures.

@@

The only non-standard feature of the typing rules is the priority annotations. Priorities are based on \emph{obligations/capabilities} used by Kobayashi~\cite{kobayashi06}, and simplified to single priorities following Padovani~\cite{padovani14}. The integration of priorities into GV is adapted from Padovani and Novara~\cite{padovaninovara15}. Paraphrasing Dardha and Gay~\cite{dardhagay18}, priorities obey the following two laws:
\begin{enumerate}
  \item an action with priority $\cs{o}$ happens before an action with priority $\cs{o+n}$;
  \item communication requires \emph{equal} priorities for the dual actions.
\end{enumerate}

In PGV\simon{I think the rules need more explanation than this---as someone who knows the basics of priorities but is not an expert, I found it difficult to understand the T-Lam rule for example}, sequents are annotated by a priority bound $\cs{p}$, which is the \emph{upper bound} on the priority of the communication actions performed while evaluating the term. The \emph{lower bound} is obtained from the typing environment. Terms with sequential evaluation, such as \LabTirName{T-LetUnit}, have side conditions to ensure deadlock freedom. For instance, the side condition for typing $\tm{\letunit{M}{N}}$ requires that $\tm{M}$ has finished communicating before $\tm{N}$ starts communicating.

Configurations are typed by a judgement $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$, which states that ``a configuration $\tm{\conf{C}}$ with flag $\phi$ is well-typed under typing environment $\ty{\Gamma}$''\simon{I found it surprising that configuration typing doesn't need priority annotations. Can you give an intuitition why?}. Configuration typing is based on the standard typing for GV. Terms are embedded either as main or as child threads. The priority bound from the term typing is discarded, as configurations contain no further sequential actions. Main threads are allowed to return a value, whereas child threads are required to return the unit value. Sequents are annotated with a flag $\phi$, which ensure that there is at most one main thread.

@@

Secondly, we simplify typing for parallel composition. In GV, each parallel composition must split \emph{exactly one} channel of the channel pseudo-type $\ty{S^\sharp}$ into two endpoints of type $\ty{S}$ and $\ty{\co{S}}$ across the parallel composition. Consequently, associativity of parallel composition does not preserve typing. Unfortunately, splitting exactly one channel across a parallel composition in GV plays a crucial role in guaranteeing deadlock freedom. In PGV, we guarantee deadlock freedom using priorities, which removes the channel pseudo-type $\ty{S^\sharp}$, simplifies typing for parallel composition, and restores type preservation for the structural congruence.

@@

\subsubsection*{Subject reduction}
Unlike with previous versions of GV, structural congruence, term reduction, and configuration reduction are all type preserving, though it is not entirely trivial to show.
